Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length within leaves?

### 2023-11-13 ms/ms.Rmd
small revisions to ms before submission

title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?

@@

  - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment. These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces. This feature is considered an adaptation to alleviate CO2 diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved. Plants can also reduce CO2 limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution and patterning, and size. A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous plants, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species. Here we use novel spatial statistics based on simulations and photosynthesis modelling to test hypotheses about how amphistomatous plants may optimize CO2 limitations in the model angiosperm Arabidopsis thaliana grown in different light environments. We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surfaces are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule (~1%) across the range of feasible traitspace. However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO2, suggesting that plants may optimize CO2 diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing. We discuss the myriad developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination. These findings contribute to our understanding of amphistomatous plants' exploration of the leaf anatomy adaptive landscape."

@@

 - \doublespacing

@@

Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017]. Therefore leaves should develop enough stomata to adequately supply CO$_2$ to chloroplasts, but not overinvest. A widespread hypothesis in plant ecophysiology is that natural selection optimizes traits like stomatal size, density, and distribution to maximize carbon gain relative to any costs in a given environmental context. In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal. However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e. the selection coefficient is less than drift barrier [@sung_drift-barrier_2012]).

@@

No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface. Rather, stomatal development is a dynamic process that must be plastic to environmental ques. Leaves develop based on short and long distance signalling pathways which relay information about incoming light, humidity, temperature, and surrounding stomata to developing leaf tissues [@pillitteri_mechanisms_2012]. Our results showing an intermediate level of dispersion in stomatal spacing may be best explained by these developmental pathways which ensure the proper spacing of stomata, with an added random effect brought about by the necessity for plasticity in stomatal development (\autoref{fig:single-surface}). However, any nonideal randomness in stomatal spacing may offset by simultaneous and coordinated development of the IAS [@baillie_developmental_2020]. The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf tissues to negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a third dimension. Our gas exchange model demonstrates little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}). Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e. the drift-barrier). We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal pattern. Any benefit garnered by ideal stomatal spacing may be additionally offset by a cost to developmental flexibility in variable environments [@pillitteri_mechanisms_2012; @baillie_developmental_2020]. Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant. As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion. Such flexibility in mature stomate size is an alternate peak in the adaptive landscape which cannot be assessed with our simplistic model assumptions. In any case, there is no evidence for coordinated development of both leaf surfaces, and very little theoretical benefit to photosynthesis, except in marginal circumstances which are exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

Our results suggest that after optimizing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization. Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution. The possibility that ideal stomatal spacing is not the "tallest" fitness peak must also be explored, as stomate size is demonstrated in this study to covary with mesophyll volume supplied with CO$_2$. This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental ques may change rapidly. The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal. In these cases, the theoretical benefits of ideal stomatal spacing are further diminished. Future exploration of these competing hypotheses would require more advanced modelling, additional exploration of IAS space development and its effects on gas exchange, both real and modeled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings. Despite these additional considerations, this paper represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.

>>>>

title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length within leaves?

@@

  - "Photosynthesis is co-limited by multiple factors depending on the plant and its environment. These include biochemical rate limitations, internal and external water potentials, temperature, irradiance, and carbon dioxide (CO$_2$). Amphistomatous leaves have stomata on both abaxial and adaxial leaf surfaces. This feature is considered an adaptation to alleviate CO$_2$ diffusion limitations in productive environments where other factors are not limiting as the diffusion path length from stomate to chloroplast is effectively halved. Plants can also reduce CO$_2$ limitations through other aspects of optimal stomatal anatomy: stomatal density, distribution, patterning, and size. A number of studies have demonstrated that stomata are overdispersed on a single leaf surface; however, much less is known about stomatal anatomy in amphistomatous leaves, especially the coordination between leaf surfaces, despite their prevelance in nature and near ubiquity among crop species. Here we use novel spatial statistics based on simulations and photosynthesis modeling to test hypotheses about how amphistomatous plants may optimize CO$_2$ limitations in the model angiosperm *Arabidopsis thaliana* grown in different light environments. We find that 1) stomata are overdispersed, but not ideally dispersed, on both leaf surfaces across all light treatments; 2) abaxial and adaxial leaf surface patterning are independent; and 3) the theoretical improvements to photosynthesis from abaxial-adaxial stomatal coordination are miniscule ($\\ll 1$%) across the range of feasible parameter space. However, we also find that 4) stomatal size is correlated with the mesophyll volume that it supplies with CO$_2$, suggesting that plants may optimize CO$_2$ diffusion limitations through alternative pathways other than ideal, uniform stomatal spacing. We discuss the developmental, physical, and evolutionary constraits which may prohibit plants from reaching the theoretical adaptive peak of uniform stomatal spacing and inter-surface stomatal coordination. These findings contribute to our understanding of variation in the anatomy of amphistomatous leaves."

@@

 - \onehalfspacing

@@

Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017]. Therefore leaves should develop enough stomata to adequately supply CO$_2$ to chloroplasts, but not overinvest. A widespread hypothesis in plant ecophysiology is that natural selection optimizes traits like stomatal size, density, and distribution to maximize carbon gain relative to any costs in a given environmental context. In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal. However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e. the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012).

@@

No developmental pathway exists to ensure an idealized placement of stomata on the leaf surface. Rather, stomatal development is a dynamic process that must be plastic to environmental cues. Leaves develop based on short and long distance signalling pathways which relay information about incoming light, humidity, temperature, and surrounding stomata to developing leaf tissues [@pillitteri_mechanisms_2012]. Our results showing an intermediate level of dispersion in stomatal spacing may be best explained by these developmental pathways which ensure the proper spacing of stomata, with an added random effect brought about by the necessity for plasticity in stomatal development (\autoref{fig:single-surface}). However, any nonideal randomness in stomatal spacing may offset by simultaneous and coordinated development of the IAS [@baillie_developmental_2020]. The fact that stomata which supply a greater mesophyll volume tend to be larger suggests that plants may use coordinated development of multiple leaf tissues to negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a third dimension. Our gas exchange model demonstrates little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}). Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e. the drift-barrier). We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal pattern. Any benefit garnered by ideal stomatal spacing may be additionally offset by a cost to developmental flexibility in variable environments [@pillitteri_mechanisms_2012; @baillie_developmental_2020]. Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant. As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion. In any case, there is no evidence for coordinated development of both leaf surfaces, and very little theoretical benefit to photosynthesis, except in marginal circumstances which are exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching a theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modeling (with a more complex set of assumptions including vein density and IAS structure) should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

Our results suggest that after optimizing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization. Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution. The possibility that ideal stomatal spacing is not the "tallest" fitness peak must also be explored, as stomate size is demonstrated in this study to covary with mesophyll volume supplied with CO$_2$. This may be especially true in highly variable environments or in large tree species with sun and shade leaves where developmental cues may change rapidly. The temporal component, not considered here, could also have significant implications, as CO$_2$ may only be limiting to photosynthesis during short periods when all other conditions are ideal. In these cases, the theoretical benefits of ideal stomatal spacing are further diminished. Future exploration of these competing hypotheses would require more advanced modeling, additional exploration of IAS space development and its effects on gas exchange, both real and modeled, and knowledge about how often the species of interest is CO$_2$ limited across of range of natural settings. Despite these additional considerations, this study represents an important contribution to understanding the potential drivers of and limitations to stomatal anatomy in amphistomatous plants.

### 2023-11-13 ms/ms.tex
small revisions to ms before submission

% one column
\onecolumn

@@

\hypersetup{
  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length with leaves of Arabidopsis thaliana?},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Pandoc header-include feature
\usepackage{setspace}
\doublespacing
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
\usepackage{newunicodechar,graphicx}
\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
\newunicodechar{ʻ}{\okina}
% Pandoc header-include feature
\usepackage{booktabs}

@@

\title[]{Does stomatal patterning in amphistomatous leaves minimize the
CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis
thaliana}?}

@@

\abstract{
Photosynthesis is co-limited by multiple factors depending on the plant
and its environment. These include biochemical rate limitations,
internal and external water potentials, temperature, irradiance, and
carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial
and adaxial leaf surfaces. This feature is considered an adaptation to
alleviate CO2 diffusion limitations in productive environments where
other factors are not limiting as the diffusion path length from stomate
to chloroplast is effectively halved. Plants can also reduce CO2
limitations through other aspects of optimal stomatal anatomy: stomatal
density, distribution and patterning, and size. A number of studies have
demonstrated that stomata are overdispersed on a single leaf surface;
however, much less is known about stomatal anatomy in amphistomatous
plants, especially the coordination between leaf surfaces, despite their
prevelance in nature and near ubiquity among crop species. Here we use
novel spatial statistics based on simulations and photosynthesis
modelling to test hypotheses about how amphistomatous plants may
optimize CO2 limitations in the model angiosperm Arabidopsis thaliana
grown in different light environments. We find that 1) stomata are
overdispersed, but not ideally dispersed, on both leaf surfaces across
all light treatments; 2) abaxial and adaxial leaf surfaces are
independent; and 3) the theoretical improvements to photosynthesis from
abaxial-adaxial stomatal coordination are miniscule (\textasciitilde1\%)
across the range of feasible traitspace. However, we also find that 4)
stomatal size is correlated with the mesophyll volume that it supplies
with CO2, suggesting that plants may optimize CO2 diffusion limitations
through alternative pathways other than ideal, uniform stomatal spacing.
We discuss the myriad developmental, physical, and evolutionary
constraits which may prohibit plants from reaching the theoretical
adaptive peak of uniform stomatal spacing and inter-surface stomatal
coordination. These findings contribute to our understanding of
amphistomatous plants' exploration of the leaf anatomy adaptive
landscape.}

@@

Stomata cost resources to maintain \citep{deans_optimization_2020} and
expose leaves to risks such as hydraulic failure
\citep{wang_theoretical_2020} or infection by plant pathogens
\citep{melotto_stomatal_2017}. Therefore leaves should develop enough
stomata to adequately supply CO\(_2\) to chloroplasts, but not
overinvest. A widespread hypothesis in plant ecophysiology is that
natural selection optimizes traits like stomatal size, density, and
distribution to maximize carbon gain relative to any costs in a given
environmental context. In principle, spacing stomata to minimize the
average distance between stomatal pores and chloroplasts within the
mesophyll should increase carbon gain, all else being equal. However,
reducing this distance to its absolute minimum may be constrained by
developmental processes, or the photosynthetic benefit may be too small
to be `seen' by natural selection (i.e.~the selection coefficient is
less than drift barrier \citep{sung_drift-barrier_2012}).

@@

No developmental pathway exists to ensure an idealized placement of
stomata on the leaf surface. Rather, stomatal development is a dynamic
process that must be plastic to environmental ques. Leaves develop based
on short and long distance signalling pathways which relay information
about incoming light, humidity, temperature, and surrounding stomata to
developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
showing an intermediate level of dispersion in stomatal spacing may be
best explained by these developmental pathways which ensure the proper
spacing of stomata, with an added random effect brought about by the
necessity for plasticity in stomatal development
(\autoref{fig:single-surface}). However, any nonideal randomness in
stomatal spacing may offset by simultaneous and coordinated development
of the IAS \citep{baillie_developmental_2020}. The fact that stomata
which supply a greater mesophyll volume tend to be larger suggests that
plants may use coordinated development of multiple leaf tissues to
negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a
third dimension. Our gas exchange model demonstrates little
photosynthetic gain from abaxial-adaxial stomatal coordination
(\autoref{fig:model_summary}). Even though lateral diffusion may limit
photosynthesis \citep{morison_lateral_2005}, the marginal gain from
optimally offsetting stomata is not sufficient to generate fitness
differences relative to the strength of genetic drift (i.e.~the
drift-barrier). We can similarly extrapolate that an ideal, equilateral
triangular stomatal spacing is only slightly better than a suboptimal
pattern. Any benefit garnered by ideal stomatal spacing may be
additionally offset by a cost to developmental flexibility in variable
environments
\citep{pillitteri_mechanisms_2012, baillie_developmental_2020}.
Explaining these observations as the result of weak selection is in
tension with the finding that stomatal size and zone positively covary,
which would suggest that small changes in lateral diffusion distance are
significant. As described above, the positive correlation between
stomatal size and zone may be explained by common developmental
processes rather than as an adaptation to maximize CO\(_2\) diffusion.
Such flexibility in mature stomate size is an alternate peak in the
adaptive landscape which cannot be assessed with our simplistic model
assumptions. In any case, there is no evidence for coordinated
development of both leaf surfaces, and very little theoretical benefit
to photosynthesis, except in marginal circumstances which are
exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata
are non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching a theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling (with a more complex
set of assumptions including vein density and IAS structure) should be
done to estimate the fitness gain of stomatal dispersion. Additionally,
genetic manipulation studies should attempt to create mutants with
clustered and ideally dispersed stomata for a comparison of their
photosynthetic traits. This could have extremely important implications
for maximum assimilation rates in crops as most crop species are grown
in high light where CO\(_2\) is often limiting. In drought-prone
environments, increased stomatal dispersion may increase water use
efficiency by reducing the number of stomata needed to achieve the same
internal CO\(_2\) concentration, \(C_\text{i}\).

Our results suggest that after optimizing stomatal density and having
developmental rules for spacing stomata relatively evenly, there may be
limited gains to further optimization. Therefore, developmental
constraints may be necessary to make sense of some features of stomatal
spacing and distribution. The possibility that ideal stomatal spacing is
not the ``tallest'' fitness peak must also be explored, as stomate size
is demonstrated in this study to covary with mesophyll volume supplied
with CO\(_2\). This may be especially true in highly variable
environments or in large tree species with sun and shade leaves where
developmental ques may change rapidly. The temporal component, not
considered here, could also have significant implications, as CO\(_2\)
may only be limiting to photosynthesis during short periods when all
other conditions are ideal. In these cases, the theoretical benefits of
ideal stomatal spacing are further diminished. Future exploration of
these competing hypotheses would require more advanced modelling,
additional exploration of IAS space development and its effects on gas
exchange, both real and modeled, and knowledge about how often the
species of interest is CO\(_2\) limited across of range of natural
settings. Despite these additional considerations, this paper represents
an important contribution to understanding the potential drivers of and
limitations to stomatal anatomy in amphistomatous plants.

>>>>

\usepackage{newunicodechar,graphicx}
\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
\newunicodechar{ʻ}{\okina}
% one column
\onecolumn

@@

\hypersetup{
  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length within leaves?},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Pandoc header-include feature
\usepackage{setspace}
\onehalfspacing
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
% Pandoc header-include feature
\usepackage{booktabs}

@@

\title[]{Does stomatal patterning in amphistomatous leaves minimize the
CO\(_2\) diffusion path length within leaves?}

@@

\abstract{
Photosynthesis is co-limited by multiple factors depending on the plant
and its environment. These include biochemical rate limitations,
internal and external water potentials, temperature, irradiance, and
carbon dioxide (CO\(_2\)). Amphistomatous leaves have stomata on both
abaxial and adaxial leaf surfaces. This feature is considered an
adaptation to alleviate CO\(_2\) diffusion limitations in productive
environments where other factors are not limiting as the diffusion path
length from stomate to chloroplast is effectively halved. Plants can
also reduce CO\(_2\) limitations through other aspects of optimal
stomatal anatomy: stomatal density, distribution, patterning, and size.
A number of studies have demonstrated that stomata are overdispersed on
a single leaf surface; however, much less is known about stomatal
anatomy in amphistomatous leaves, especially the coordination between
leaf surfaces, despite their prevelance in nature and near ubiquity
among crop species. Here we use novel spatial statistics based on
simulations and photosynthesis modeling to test hypotheses about how
amphistomatous plants may optimize CO\(_2\) limitations in the model
angiosperm \emph{Arabidopsis thaliana} grown in different light
environments. We find that 1) stomata are overdispersed, but not ideally
dispersed, on both leaf surfaces across all light treatments; 2) abaxial
and adaxial leaf surface patterning are independent; and 3) the
theoretical improvements to photosynthesis from abaxial-adaxial stomatal
coordination are miniscule (\(\ll 1\)\%) across the range of feasible
parameter space. However, we also find that 4) stomatal size is
correlated with the mesophyll volume that it supplies with CO\(_2\),
suggesting that plants may optimize CO\(_2\) diffusion limitations
through alternative pathways other than ideal, uniform stomatal spacing.
We discuss the developmental, physical, and evolutionary constraits
which may prohibit plants from reaching the theoretical adaptive peak of
uniform stomatal spacing and inter-surface stomatal coordination. These
findings contribute to our understanding of variation in the anatomy of
amphistomatous leaves.}

@@

Stomata cost resources to maintain \citep{deans_optimization_2020} and
expose leaves to risks such as hydraulic failure
\citep{wang_theoretical_2020} or infection by plant pathogens
\citep{melotto_stomatal_2017}. Therefore leaves should develop enough
stomata to adequately supply CO\(_2\) to chloroplasts, but not
overinvest. A widespread hypothesis in plant ecophysiology is that
natural selection optimizes traits like stomatal size, density, and
distribution to maximize carbon gain relative to any costs in a given
environmental context. In principle, spacing stomata to minimize the
average distance between stomatal pores and chloroplasts within the
mesophyll should increase carbon gain, all else being equal. However,
reducing this distance to its absolute minimum may be constrained by
developmental processes, or the photosynthetic benefit may be too small
to be `seen' by natural selection (i.e.~the selection coefficient is
less than drift barrier \emph{sensu} \citet{sung_drift-barrier_2012}).

@@

No developmental pathway exists to ensure an idealized placement of
stomata on the leaf surface. Rather, stomatal development is a dynamic
process that must be plastic to environmental cues. Leaves develop based
on short and long distance signalling pathways which relay information
about incoming light, humidity, temperature, and surrounding stomata to
developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
showing an intermediate level of dispersion in stomatal spacing may be
best explained by these developmental pathways which ensure the proper
spacing of stomata, with an added random effect brought about by the
necessity for plasticity in stomatal development
(\autoref{fig:single-surface}). However, any nonideal randomness in
stomatal spacing may offset by simultaneous and coordinated development
of the IAS \citep{baillie_developmental_2020}. The fact that stomata
which supply a greater mesophyll volume tend to be larger suggests that
plants may use coordinated development of multiple leaf tissues to
negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a
third dimension. Our gas exchange model demonstrates little
photosynthetic gain from abaxial-adaxial stomatal coordination
(\autoref{fig:model_summary}). Even though lateral diffusion may limit
photosynthesis \citep{morison_lateral_2005}, the marginal gain from
optimally offsetting stomata is not sufficient to generate fitness
differences relative to the strength of genetic drift (i.e.~the
drift-barrier). We can similarly extrapolate that an ideal, equilateral
triangular stomatal spacing is only slightly better than a suboptimal
pattern. Any benefit garnered by ideal stomatal spacing may be
additionally offset by a cost to developmental flexibility in variable
environments
\citep{pillitteri_mechanisms_2012, baillie_developmental_2020}.
Explaining these observations as the result of weak selection is in
tension with the finding that stomatal size and zone positively covary,
which would suggest that small changes in lateral diffusion distance are
significant. As described above, the positive correlation between
stomatal size and zone may be explained by common developmental
processes rather than as an adaptation to maximize CO\(_2\) diffusion.
In any case, there is no evidence for coordinated development of both
leaf surfaces, and very little theoretical benefit to photosynthesis,
except in marginal circumstances which are exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata
are non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching a theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modeling (with a more complex
set of assumptions including vein density and IAS structure) should be
done to estimate the fitness gain of stomatal dispersion. Additionally,
genetic manipulation studies should attempt to create mutants with
clustered and ideally dispersed stomata for a comparison of their
photosynthetic traits. This could have extremely important implications
for maximum assimilation rates in crops as most crop species are grown
in high light where CO\(_2\) is often limiting. In drought-prone
environments, increased stomatal dispersion may increase water use
efficiency by reducing the number of stomata needed to achieve the same
internal CO\(_2\) concentration, \(C_\text{i}\).

Our results suggest that after optimizing stomatal density and having
developmental rules for spacing stomata relatively evenly, there may be
limited gains to further optimization. Therefore, developmental
constraints may be necessary to make sense of some features of stomatal
spacing and distribution. The possibility that ideal stomatal spacing is
not the ``tallest'' fitness peak must also be explored, as stomate size
is demonstrated in this study to covary with mesophyll volume supplied
with CO\(_2\). This may be especially true in highly variable
environments or in large tree species with sun and shade leaves where
developmental cues may change rapidly. The temporal component, not
considered here, could also have significant implications, as CO\(_2\)
may only be limiting to photosynthesis during short periods when all
other conditions are ideal. In these cases, the theoretical benefits of
ideal stomatal spacing are further diminished. Future exploration of
these competing hypotheses would require more advanced modeling,
additional exploration of IAS space development and its effects on gas
exchange, both real and modeled, and knowledge about how often the
species of interest is CO\(_2\) limited across of range of natural
settings. Despite these additional considerations, this study represents
an important contribution to understanding the potential drivers of and
limitations to stomatal anatomy in amphistomatous plants.
